Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2019-11-02 manuscript/climate_sensitivity_SI.Rmd
added Table S3- drought stats

```{r, Table S3, eval=TRUE, echo=FALSE, message=FALSE}
tableS3 <- read.csv("tables_figures/tableS3_candidate_traits.csv", stringsAsFactors = FALSE)
kable(tableS3, booktabs=TRUE, format = "latex", caption = "Table S3: Candidate variables for best model") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
```

@@

```{r Table S4, eval=TRUE, echo=FALSE, message=FALSE}
tableS5 <- read.csv("tables_figures/tableS5_tested_traits_height.csv", stringsAsFactors = FALSE)

>>>>

```{r, Table S3, eval=TRUE, echo=FALSE, message=FALSE}
tableS3 <- read.csv("tables_figures/drought_stats.csv", stringsAsFactors = FALSE)
kable(tableS3, booktabs=TRUE, format = "latex", caption = "Table S3: Palmer drought severity index (PDSI) by month for focal droughts and other years referenced in the manuscript") %>%
  pack_rows("focal droughts", 1, 4, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 5, 8, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 9, 12, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("others", 13, 16, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 17, 20, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 21, 24, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
```

@@

```{r, Table S4, eval=TRUE, echo=FALSE, message=FALSE}
tableS4 <- read.csv("tables_figures/tableS3_candidate_traits.csv", stringsAsFactors = FALSE)
kable(tableS4, booktabs=TRUE, format = "latex", caption = "Table S4: Candidate variables for best model") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE, latex_options = "hold_position")
```

```{r Table S5, eval=TRUE, echo=FALSE, message=FALSE}
tableS5 <- read.csv("tables_figures/tableS5_tested_traits_height.csv", stringsAsFactors = FALSE)

### 2019-11-02 manuscript/climate_sensitivity_draft.Rmd
added Table S3- drought stats

Second, for each drought year and for all years combined, we determined the best multivariate models for predicting $R. All variables with dAICc > 1.0 in any model (drought years combined or any individual drought) were used as candidate variables for each scenario's best model (Table S3). We compared models with all possible combinations of candidate variables and identified the full set of models within dAICc=2 of the top model (that with lowest AICc). These were counted as the "top multivariate models". When a variable appeared in all of these top models and the sign of the coefficient was consistent across models, this was counted as support for/ rejection of the associated prediction by the multivariate models. If the variable appeared in only some of the models, we considered this partial support.

@@

In the 60-year period between 1950 and 2009, there were three droughts that met our criteria of anomalously dry climatic conditions coupled with substantial reductions in tree growth for at least some portions of the community: 1966, 1977, and 1999 (Fig. 1). 
We excluded one year (1991) meeting the growth reduction criteria (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not among the top ten driest of the study period in terms of May-August [PET-PRE] or PDSI. Rather, the severity of growth reduction may be explained by defoliation by gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which most stronly impacted *Quercus* spp. (*Cite Shenandoah paper, if accepted*). 
Climatically, these droughts included three of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August, which are the months to which annual tree growth at this site is most sensitive [@helcoski_growing_2019]. Specifically, 1966, 1977, and 1999 had mean MJJA PET-PRE of 83.37, 86.97, and 80 mm mo<sup>-1</sup>, respectively. The years 1964 and 2007 also ranked in the top five driest (PET-PRE =83.87 and 82.13 mm mo<sup>-1</sup>), but *were not among the lowest in terms of PDSI and were not identified as a pointer yeasr.*

@@

We partially support *H2.1*: Species' hydraulic traits --$XP$, $PLA_{dry}$, and $\pi_{tlp}$--were sometimes useful in explaining variation in drought responses, whereas $LMA$ and $WD$ were not (Tables 1,4,5). Specifically, $LMA$ and $WD$ never significantly associated with $R$ in the univariate models (all dAIC $\le$ 0.22; Table 4), and therefore these were excluded as candidate variables for the full multivariate models (Table S3). In contrast, $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained at least modest amounts of variation (dAIC > 1.0) in at least one drought (Table 4). Of these, $PLA_{dry}$ was the strongest predictor, with consistently negative coefficients across all droughts. $\pi_{tlp}$ never came out as significant (dAIC $\ge$ 2) in the univariate models, but had a consistently negative coefficient (Table 4). Whereas ring-porous species had highest $R$ overall and in the 1966 and 1999 droughts, they had lower $R$ in 1977. Results were similar in the context of multivariate models (Table 5), except that $\pi_{tlp}$ had a positive coefficient in the 1966 models in which it was included.

We reject *H2.2*, finding no evidence that taller trees tend to have traits associated with greater drought resistance. In part because of the large sample size (**n=#** trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha), there were very significant (p<0.0001) correlations of $H$ with all species' traits (see Table S4). However, the correlation only matched the predicted direction (*i.e.*, more drought-resistant traits associated to taller trees) in the cases of $WD$ $LMA$, which were not useful predictors of $R$. Furthermore, although correlations were statistically signifcant, trait variation within each height class overwhelmed any vertical trends (Fig. 2e-f).

>>>>

Second, for each drought year and for all years combined, we determined the best multivariate models for predicting $R. All variables with dAICc > 1.0 in any model (drought years combined or any individual drought) were used as candidate variables for each scenario's best model (Table S4). We compared models with all possible combinations of candidate variables and identified the full set of models within dAICc=2 of the top model (that with lowest AICc). These were counted as the "top multivariate models". When a variable appeared in all of these top models and the sign of the coefficient was consistent across models, this was counted as support for/ rejection of the associated prediction by the multivariate models. If the variable appeared in only some of the models, we considered this partial support.

@@

In the 60-year period between 1950 and 2009, there were three droughts that met our criteria of anomalously dry climatic conditions coupled with substantial reductions in tree growth for at least some portions of the community: 1966, 1977, and 1999 (Fig. 1). 
We excluded one year (1991) meeting the growth reduction criteria (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not among the driest in terms of May-August [PET-PRE] or PDSI (Table S3). Rather, the severity of growth reduction may be explained by defoliation by gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which most stronly impacted *Quercus* spp. (*Cite Shenandoah paper, if accepted*). 
Climatically, these droughts included three of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August, which are the months to which annual tree growth at this site is most sensitive [@helcoski_growing_2019]. Specifically, 1966, 1977, and 1999 had mean MJJA PET-PRE of 83.37, 86.97, and 80 mm mo<sup>-1</sup>, respectively. The years 1964 and 2007 also ranked in the top five driest (PET-PRE =83.87 and 82.13 mm mo<sup>-1</sup>), but were not among the lowest in terms of PDSI and were not identified as a pointer yeasr (Table S3)

@@

We partially support *H2.1*: Species' hydraulic traits --$XP$, $PLA_{dry}$, and $\pi_{tlp}$--were sometimes useful in explaining variation in drought responses, whereas $LMA$ and $WD$ were not (Tables 1,4,5). Specifically, $LMA$ and $WD$ never significantly associated with $R$ in the univariate models (all dAIC $\le$ 0.22; Table 4), and therefore these were excluded as candidate variables for the full multivariate models (Table S4). In contrast, $XP$, $PLA_{dry}$, and $\pi_{tlp}$ all explained at least modest amounts of variation (dAIC > 1.0) in at least one drought (Table 4). Of these, $PLA_{dry}$ was the strongest predictor, with consistently negative coefficients across all droughts. $\pi_{tlp}$ never came out as significant (dAIC $\ge$ 2) in the univariate models, but had a consistently negative coefficient (Table 4). Whereas ring-porous species had highest $R$ overall and in the 1966 and 1999 droughts, they had lower $R$ in 1977. Results were similar in the context of multivariate models (Table 5), except that $\pi_{tlp}$ had a positive coefficient in the 1966 models in which it was included.

We reject *H2.2*, finding no evidence that taller trees tend to have traits associated with greater drought resistance. In part because of the large sample size (**n=#** trees--all individuals of our 12 focal species $\ge$ 10 cm in 25.6 ha), there were very significant (p<0.0001) correlations of $H$ with all species' traits (see Table S5). However, the correlation only matched the predicted direction (*i.e.*, more drought-resistant traits associated to taller trees) in the cases of $WD$ $LMA$, which were not useful predictors of $R$. Furthermore, although correlations were statistically signifcant, trait variation within each height class overwhelmed any vertical trends (Fig. 2e-f).
